Screening of airborne radioxenon measurements for Nuclear-Test-Ban Treaty verification

### 2022-10-26 paper/paper.tex
Moved anomaly detection chapter before conclusions and added it to the outline Minor corrections and clarifications

\subsection{Outline}
The above assumes that the xenon emission is constant for every source. We adapt this model in \cref{sec: model} so that daily emissions are taken into account. In \cref{sec: nondetections}, we discuss the use of a transformation from \cite{deMeutter2022uncertainty} to model the accuracy in a non-detection event.
In \cref{sec: optimisation}, we explain a method to fit the extended model to the measured concentration data. Given the time constraints of this project, the optimality criterion was deliberately kept simple. Numerical data in \cref{sec: numerical results} show that the average prediction error can be decreased. We conclude in \cref{sec: further work} with suggestions to make the model more realistic and improve computational aspects.

@@

\Cref{eq: initial model} can be formulated as follows:
\begin{align}
    \label{eq: initial model concise} 
\begin{bmatrix}
    \hat{y}_1 \\ 
    \vdots \\
    \hat{y}_{n}
\end{bmatrix}
&=
\begin{bmatrix}
    M_{1,(1,1)} & M_{1,(1,2)} & \cdots & M_{1,(m,15)} \\
    \vdots & \vdots & & \vdots \\
    M_{n,(1,1)} & M_{n,(1,2)} & \cdots & M_{n,(m,15)} \\
\end{bmatrix}
\begin{bmatrix}
    s_1^{-1} x_1\\
    \vdots \\
    s_1^{-1} x_1 \\
    s_2^{-1} x_2 \\
    \vdots \\
    s_n^{-1} x_m
\end{bmatrix} 
\\
&=
\begin{bmatrix}
    M_{1,(1,1)} & M_{1,(1,2)} & \cdots & M_{1,(n,15)} \\
    \vdots & \vdots & & \vdots \\
    M_{m,(1,1)} & M_{m,(1,2)} & \cdots & M_{m,(n,15)} \\
\end{bmatrix}
\left( 
     S^{-1}
     \begin{bmatrix}
        x_1 \\
        \vdots \\
        x_n
     \end{bmatrix}
     \otimes \vb{1}_{15}
\right)
\nonumber
\end{align}
where $S = \mathrm{diag}([s_1,\dots,s_m])$.
In this section, we present a different interpretation. Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x} := S^{-1} x \otimes \vb{1}_{T}$ where $T = 365 + 15 = 380$. This is a vector of length $mT$ that gives the emission from the most recent day to $T$ days in the past for each source. This is assumed to be constant for all days. Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is a constant vector of length $T$.

@@

\begin{remark}
    As an alternative to using $\tilde{x} = S^{-1} x \otimes \vb{1}_T$, one could use the permuted expression $\vb{1}_{T} \otimes (S^{-1} x)$. This is a vector containing all emissions on day 1, followed by all emissions on day 2, etc. A similar permutation of $y$ could be chosen. These permutations can be compensated by the inverse permutation on the columns and rows of $M_{shift}$, respectively. The resulting model is essentially identical to \cref{eq: initial model 365 days} but may have advantages when it comes to implementation in software. We will not consider this.
\end{remark}

\subsection{Model with time-dependent correction to the emissions}
Daily emissions can be modelled by multiplying the average emission of source $j$ by some scaling factor $w_{j,t}$. This corresponds to the 
transformation
$$
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
\mapsto
\begin{bmatrix}
    w_{1,1} s_1^{-1} x_1 \\
    \vdots \\
    w_{1,T} s_1^{-1} x_1 \\ 
    w_{2,1} s_2^{-1} x_2 \\
    \vdots \\ 
    w_{m, T} s_m^{-1} x_m
\end{bmatrix}
$$
which can be written concisely as $W\tilde{x}$ where $W = \mathrm{diag}([w_{1,1}, w_{1,2},\dots,w_{m,T}])$.
Using this weighted vector of emissions instead of $\tilde{x}$ in \cref{eq: initial model 365 days} gives
\begin{equation}
    \label{eq: time-evolving model concise}
\hat{y} = M_{shift} W \tilde{x}
.\end{equation}
In the next section, we present a method to fit the unknown factors $W$ to the data.

@@

Sometimes, the concentration at a measurement station is below the minimum detectable concentration (MDC). The corresponding measurements are $y_i = 0$.
To account for non-detections, we apply the transformation $h$ from \cite{deMeutter2022uncertainty}. This is given by the continuously differentiable function
$$
h(y) = \begin{cases}
    \frac{y^2}{4MDC} + MDC &\quad \text{if } y \le 2MDC \\
    y &\quad \text{else}
\end{cases}
.$$
This transformation maps the interval $[0, MDC]$ onto $[MDC, 2MDC]$. For observations that are significantly larger than $MDC$, this is approximately the identity. If $y$ and the simulated value $\hat{y}$ are both larger than $2MDC$, then $h(\hat{y}) / h(y) = \hat{y} / y$. If $y = 0$ because the concentration is not detectable, then we want a simulated value $\hat{y}$ such that $h(\hat{y}) \approx MDC$. In both cases, we can use the value of $h(\hat{y}) / h(y)$ as an indication of the accuracy of the simulation.

@@

The second criterion is the ratio between the daily emissions and the average emission for each source, i.e. the factors $W$. In a similar vein as the above, the size of $W$ can be measured as $\norm{\log W}^2$ where $\norm{\cdot}$ is the Frobenius norm\footnote{Since $W$ is a diagonal matrix, the matrix logarithm $\log W$ is simply the diagonal matrix with $\log w_{1,1}, \log w_{1,2}, \dots$ on the diagonal. Hence, $\norm{\log W}^2 = \log^2 w_{1,1} + \log^2 w_{1,2} + \dots$}.

@@

The previous metric only takes into account the simulated observations within a small area. Therefore it cannot tell us whether some general bias is present. To investigate this, we use the geometric mean bias. The more biased the ratio is, the further this value will lie from $1$. This is because it defines some mean ratio of actual observations and simulated ones.
%Nitpick Thomas: err, I did not actually check the formulae for the metrics closely during the week. It seems slightly weird taking the exponential, as we mostly care about the scalings (in log space). It might complicate the explanation: "needs to be close to 1 (in some logarithmic way)". Also, why do we have this minus sign in front?
\begin{equation}
	\text{GM}:=e^{-E(ln(\hat{y}/y))}
\end{equation}

@@

@@

\section{Anomaly detection}

One of the main challenges of applying traditional anomaly detection methods on the time-series data of an observation station, is accommodating for weather. Winds will change the course of an emission, causing different observation stations to pick up their signature. Therefore treating stations individually to try to learn their normal behavior would not lead to fruitful results. At the same time, it is also difficult to train more complex models that handle data collectively from all stations, since there is not enough historic data, and only a few observation stations are actively collecting data throughout the entire year.\\

>>>>

\subsection{Outline}
The model above assumes that the xenon emission is constant for every source. We adapt this model in \cref{sec: model} so that daily emissions are taken into account. In \cref{sec: nondetections}, we discuss the use of a transformation from \cite{deMeutter2022uncertainty} to model the accuracy in a non-detection event.
In \cref{sec: optimisation}, we explain a method to fit the extended model to the measured concentration data. Given the time constraints of this project, the optimality criterion was deliberately kept simple. Numerical data in \cref{sec: numerical results} show that the average prediction error can be decreased. In \cref{sec:anomaly detection} we discuss how anomaly detection could be applied on the resulting time-corrected emissions.
We conclude in \cref{sec: further work} with suggestions to make the model more realistic, and improve computational aspects.

@@

\Cref{eq: initial model} can be formulated as follows:
\begin{align}
    \label{eq: initial model concise} 
\begin{bmatrix}
    \hat{y}_1 \\ 
    \vdots \\
    \hat{y}_{n}
\end{bmatrix}
&=
\begin{bmatrix}
    M_{1,(1,1)} & M_{1,(1,2)} & \cdots & M_{1,(m,15)} \\
    \vdots & \vdots & & \vdots \\
    M_{n,(1,1)} & M_{n,(1,2)} & \cdots & M_{n,(m,15)} \\
\end{bmatrix}
\begin{bmatrix}
    s_1^{-1} x_1\\
    \vdots \\
    s_1^{-1} x_1 \\
    s_2^{-1} x_2 \\
    \vdots \\
    s_n^{-1} x_m
\end{bmatrix} 
\\
&=
\begin{bmatrix}
    M_{1,(1,1)} & M_{1,(1,2)} & \cdots & M_{1,(n,15)} \\
    \vdots & \vdots & & \vdots \\
    M_{m,(1,1)} & M_{m,(1,2)} & \cdots & M_{m,(n,15)} \\
\end{bmatrix}
\left( 
     S^{-1}
     \begin{bmatrix}
        x_1 \\
        \vdots \\
        x_n
     \end{bmatrix}
     \otimes \vb{1}_{15}
\right)
\nonumber
\end{align}
where $S = \mathrm{diag}([s_1,\dots,s_m])$.
In this section, we present a different interpretation. Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x} := S^{-1} x \otimes \vb{1}_{T}$ where $T = 365 + 15 = 380$. This is a vector of length $mT$ that gives the emission from the most recent day to $T$ days in the past for each source. This is assumed to be constant for all days (for now). Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is a constant vector of length $T$.

@@

\begin{remark}
    As an alternative to using $\tilde{x} = S^{-1} x \otimes \vb{1}_T$, one could use the permuted expression $\vb{1}_{T} \otimes (S^{-1} x)$. This is a vector containing all emissions on day 1, followed by all emissions on day 2, etc. A similar permutation of $y$ could be chosen. These permutations can be compensated by the inverse permutation on the columns and rows of $M_{shift}$, respectively. The resulting model is essentially identical to \cref{eq: initial model 365 days} but may have advantages when it comes to implementation in software (e.g. sparsity structure). We will not consider this.
\end{remark}

\subsection{Model with time-dependent correction to the emissions}
Daily emission variations can be modelled by multiplying the average emission of source $j$ by some scaling factor $w_{j,t}$. This corresponds to the 
transformation
$$
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
\mapsto
\begin{bmatrix}
    w_{1,1} s_1^{-1} x_1 \\
    \vdots \\
    w_{1,T} s_1^{-1} x_1 \\ 
    w_{2,1} s_2^{-1} x_2 \\
    \vdots \\ 
    w_{m, T} s_m^{-1} x_m
\end{bmatrix}
$$
which can be written concisely as $W\tilde{x}$ where $W = \mathrm{diag}([w_{1,1}, w_{1,2},\dots,w_{m,T}])$.
Using this weighted vector of emissions instead of $\tilde{x}$ in \cref{eq: initial model 365 days} gives
\begin{equation}
    \label{eq: time-evolving model concise}
\hat{y} = M_{shift} W \tilde{x}
.\end{equation}
In the next section, we present a method to fit the unknown factors $w_{j,t}$ to the data.

@@

Sometimes, the concentration at a measurement station is below the Minimum Detectable Concentration (MDC). The corresponding measurements are $y_i = 0$.
To account for non-detections, we apply the transformation $h$ from \cite{deMeutter2022uncertainty}. This is given by the continuously differentiable function
$$
h(y) = \begin{cases}
    \frac{y^2}{4MDC} + MDC &\quad \text{if } y \le 2MDC \\
    y &\quad \text{else}
\end{cases}
.$$
This transformation maps the interval $[0, MDC]$ onto $[MDC, 2MDC]$. For observations that are significantly larger than $MDC$, this is approximately the identity. If $y$ and the simulated value $\hat{y}$ are both larger than $2MDC$, then $h(\hat{y}) / h(y) = \hat{y} / y$. If $y = 0$ because the concentration is not detectable, then we want a simulated value $\hat{y}$ such that $h(\hat{y}) \approx MDC$. In both cases, we can use the value of $h(\hat{y}) / h(y)$ as an indication of the accuracy of the simulation.

@@

The second criterion is the ratio between the daily emissions and the average emission for each source, i.e. the factors $W$. This will act as a regularisation term that keeps the daily emissions from varying wildly. In a similar vein as the above, the size of $W$ can be measured as $\norm{\log W}^2$ where $\norm{\cdot}$ is the Frobenius norm\footnote{Since $W$ is a diagonal matrix, the matrix logarithm $\log W$ is simply the diagonal matrix with $\log w_{1,1}, \log w_{1,2}, \dots$ on the diagonal. Hence, $\norm{\log W}^2 = \log^2 w_{1,1} + \log^2 w_{1,2} + \dots$}.

@@

The previous metric only takes into account the statistical spread of simulated observations within a small area. Therefore it cannot tell us whether some general bias is present. To investigate this, we use the geometric mean bias. The more biased the ratio is, the further this value will lie from $1$. This is because it defines some mean ratio of actual observations and simulated ones.
%Nitpick Thomas: err, I did not actually check the formulae for the metrics closely during the week. It seems slightly weird taking the exponential, as we mostly care about the scalings (in log space). It might complicate the explanation: "needs to be close to 1 (in some logarithmic way)". Also, why do we have this minus sign in front?
\begin{equation}
	\text{GM}:=e^{-E(ln(\hat{y}/y))}
\end{equation}

@@

\section{Anomaly detection}
\label{sec:anomaly detection}
One of the main challenges of applying traditional anomaly detection methods on the time-series data of an observation station, is accommodating for weather. Winds will change the course of an emission, causing different observation stations to pick up their signature. Therefore treating stations individually to try to learn their normal behavior would not lead to fruitful results. At the same time, it is also difficult to train more complex models that handle data collectively from all stations, since there is not enough historic data, and only a few observation stations are actively collecting data throughout the entire year.\\

@@
